---
title: "hw03-gapminder"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load}
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("gapminder"))
```


### Task 1:
To get the max and min of gdp per continent:

```{r max_min_gdp}
summary_gdp <- gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  group_by(continent) %>% 
  summarize(max_gdp = max(gdp),
            min_gdp = min(gdp))

knitr::kable(summary_gdp)
```

```{r plot_max_gdp}
summary_gdp %>% 
  ggplot(aes(continent, max_gdp, fill = continent)) +
  geom_bar(stat = "identity") # height of a bar will be equal to the value in df
```

Africa has the lowest maximum gdp over the years, followed by Oceania. Americas have the highest maximum gdp.

```{r plot_min_gdp}
summary_gdp %>% 
  ggplot(aes(continent, min_gdp, fill = continent)) +
  geom_bar(stat = "identity") # height of a bar will be equal to the value in df
```

Oceania has a much higher minimum gdp compared to the other continents; Africa has the lowest minimum gdp.

### Task 2:
How is life expectancy changing over time on different continents?

```{r lifeExp_over_time}
lifeExp_vs_time <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mean_lifeExp = mean(lifeExp))

knitr::kable(lifeExp_vs_time)
```

```{r plot_lifeExp_vs_time}
lifeExp_vs_time %>% 
  ggplot(aes(year, mean_lifeExp, colour = continent)) +
  geom_point() +
  facet_wrap(~continent) +
  theme(legend.position="none") +  # removes the legend
  ylim(0, 100) # sets the limits for the y-axis - will be the same range 0-100 for all continents
#The first two continents don't seem to have x-axis labels; probaby because there are plots with the same axis below
```

Mean life expectancy over time is increasing in all continents. Africa has the lowest mean life expectancy; Oceania has the highest.


### Task 3:
Find countries with interesting stories:

```{r Americas_gdp_vs_lifeExp}
am_gdp <- gapminder %>%
  filter(continent == "Americas") %>%
  group_by(country) %>%
  mutate(gdp = gdpPercap*pop) %>%
  summarize(tm_gdp = mean(gdp, 0.1), # trim the mean by removing 20% of the gdp cases (top 10% and bottom 10%)
            tm_lifeExp = mean(lifeExp, 0.1)) %>% 
  arrange(tm_gdp) # from low to high

knitr::kable(am_gdp)
```

```{r plot_tmlifeExp_vs_tmgdp}
am_gdp %>%
  ggplot(aes(tm_lifeExp, tm_gdp, colour = country)) +
  geom_point()
```


United States has much bigger gdp mean over the years compared to the other countries and one of the highest life expectancies.

### Task 4:
Compute a weighted mean of life expectancy, weighting by population:

```{r wm_lifeExp}
wm_df <- gapminder %>%
  group_by(continent, year) %>% 
  summarize(wm_lifeExp = weighted.mean(lifeExp, pop))

knitr::kable(wm_df)
```

```{r plot_wm_lifeExp}
wm_df %>% 
  ggplot(aes(year, wm_lifeExp, colour = continent)) +
  geom_point() +
  facet_wrap(~continent) +
  theme(legend.position="none")   # removes the legend
```

Weighted mean of life expectancy for Africa over the years is lower compared to the other continent; Oceania still has the highest life expectancy. In Asia, there is a dip in life expectancy in the early 1960's - something that was not seen on the previous plot of lifeExp vs year.